---
title: "Car Seat Sales Forecating Using CART Regression Tree"
author: "Pranai Basava"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---
## Executive Summary

This project develops a **CART (Classification and Regression Tree)** regression model to forecast car seat sales across retail stores using demographic, pricing, and merchandising variables.

Using **5-fold cross-validation**, the model identifies key drivers of sales performance, including shelf location quality, pricing strategy, and local demographic factors.

The trained model is applied to **new, unseen store data** to generate forward-looking sales forecasts that can support:

- Inventory planning and demand forecasting  
- Store-level pricing strategy  
- Targeted advertising budget allocation  

This CART model serves as an **interpretable baseline**, establishing a foundation for future comparison with advanced models such as **Random Forests and Neural Networks**.



## 1. Project Overview

Accurately forecasting product sales is critical for inventory planning, pricing strategy, and marketing optimization.  
This project focuses on predicting **car seat sales** using store-level, demographic, and marketing-related attributes.

This notebook establishes a **baseline predictive model** using a **CART (Classification and Regression Tree)** approach.  
CART is intentionally chosen as the first model due to its interpretability and ability to capture non-linear relationships.

---

## 2. Business Problem

Retail managers need to understand:
- Which factors most strongly influence car seat sales
- How pricing, advertising, and demographics interact
- Whether interpretable models can provide actionable insights

The goal is to:
- Predict expected sales volume  
- Establish a baseline benchmark for more advanced models  
- Identify key drivers of sales performance  

---

## 3. Dataset Description

The dataset contains store-level information related to pricing, competition, demographics, and marketing spend.

### Target Variable
- **Sales**: Number of car seats sold (measured in hundreds of units)

### Predictor Variables

| Variable | Description |
|--------|------------|
| Store | Unique store identifier (excluded from modeling) |
| CompPrice | Competitor's price |
| Income | Local average income (in $000s) |
| Advertising | Advertising budget (in $000s) |
| Population | Local population size (in $000s) |
| Price | Store’s price |
| ShelfLoc | Shelf location quality (Good, Medium, Bad) |
| Age | Average local age |
| Education | Average years of education |
| Urban | Urban location (Yes/No) |
| US | Store located in US (Yes/No) |

---

## 4. Environment Setup

```{r message=FALSE, warning=FALSE}
library(caret)
library(rpart)
library(rpart.plot)
library(dplyr)
library(ggplot2)
```

## 5. Data Ingestion

```{r}
car_seat_sales <- read.csv("data/car_seat_sales.csv")

# Preview the data
head(car_seat_sales)
```

---

## 6. Data Preparation

- Remove non-predictive identifier variables  
- Let CART internally handle categorical variables  
- Preserve all relevant features for interpretability  

```{r}
library(dplyr)

# Remove Store ID
model_data <- car_seat_sales %>%
  select(-Store)

str(model_data)
```

---

## 7. Modeling Strategy

### Why CART?

CART was selected as the baseline modeling approach due to its:

- Ability to capture **non-linear relationships**
- Automatic detection of **feature interactions**
- High interpretability through decision rules
- Suitability as a benchmark before ensemble methods

### Validation Approach

To ensure robust model performance, a **5-fold cross-validation** strategy was used:

- Reduces overfitting
- Provides stable performance estimates
- Ensures generalizability beyond training data

---

## 8. Model Training – CART Regression Tree

```{r cart-training, message=FALSE, warning=FALSE}
set.seed(1234)

library(caret)

cv_method <- trainControl(
  method = "cv",
  number = 5
)

cart_model <- train(
  Sales ~ .,
  data = model_data,
  method = "rpart",
  trControl = cv_method,
  tuneLength = 100
)

cart_model
```

---

## 9. Hyperparameter Explanation (Cost Complexity – cp)

The key hyperparameter in CART is **cost complexity (cp)**.

- Controls tree depth and pruning  
- Penalizes excessive model complexity  
- Balances bias vs variance  

> A lower `cp` allows deeper trees, while a higher `cp` enforces simpler models.

The optimal `cp` is selected based on **minimum cross-validated RMSE**.

---

## 10. Model Visualization

```{r}
library(rpart.plot)

rpart.plot(cart_model$finalModel)
```

The CART regression tree reveals that **Shelf Location**, **Price**, **Advertising**, and **Competitor Pricing**
are the most influential variables affecting car seat sales.

Early splits emphasize merchandising and pricing decisions, indicating that in-store placement and competitive
pricing strategy play a dominant role in sales performance.


---

## 11. Model Performance Summary

```{r}
cart_model$results %>%
  arrange(RMSE) %>%
  head()
```

- **RMSE** – prediction error magnitude  
- **R-squared** – explained variance  

---

## 12. Generating Predictions on New Store Data

In real-world retail environments, forecasting models are often applied to **new or upcoming stores**
where historical sales data is unavailable.

To simulate this scenario, the trained CART model is applied to a new dataset containing store attributes
without observed sales values. This enables proactive demand estimation prior to store launch or product rollout.

```{r}
# Load new, unseen data
new_car_seat_sales <- read.csv("data/new_car_seat_sales_data.csv")

# Inspect the structure
str(new_car_seat_sales)
new_model_data <- new_car_seat_sales %>%
select(-Store)
```
# Predict sales for new stores
```{r}
new_predictions <- predict(cart_model, new_model_data)
```
# Combine predictions with input data
```{r}
prediction_results <- new_car_seat_sales %>%
mutate(Predicted_Sales = new_predictions)

head(prediction_results)
```

**Predicted_Sales** represents the expected number of car seats sold  
(measured in **hundreds of units**) for each store configuration.


---

## 13. Business Insights

By applying the CART model to new store-level data, decision-makers can proactively estimate expected
car seat demand before product deployment.

This enables:

- Pre-launch inventory optimization  
- Location-specific pricing adjustments  
- Targeted advertising allocation  
- Risk reduction for underperforming store profiles  

The interpretability of CART allows business stakeholders to directly understand
**why** certain stores are forecasted to perform better than others.






