---
title: "Car Seat Sales Forecasting Using Multi-Linear Regression"
author: "Pranai Basava"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

## Executive Summary

This analysis applies a **Multi-Linear Regression (MLR)** model to forecast car seat sales across retail stores using store-level demographic, pricing, and merchandising attributes.

MLR serves as a **transparent, interpretable baseline** that allows direct estimation of the influence of each predictor on sales. The model is trained using the historical dataset and evaluated using **cross-validated performance metrics** to ensure comparability with other models in this project.

The trained model is applied to **new store data** to generate forward-looking sales forecasts supporting:

- Inventory and demand planning  
- Store-level pricing strategy  
- Advertising budget optimization  
- Risk mitigation for underperforming store configurations  

This MLR model complements CART, Random Forest, and Neural Network approaches, offering **interpretability and actionable insights** into sales drivers.

---

## 1. Project Overview

Multi-Linear Regression is a **classic, interpretable regression technique** that models the relationship between multiple predictors and a continuous outcome.

In this notebook:
- **Target variable**: Sales  
- **Predictors**: Store-level pricing, advertising, demographics, merchandising  
- **Evaluation**: 5-fold cross-validation for performance benchmarking  
- **Prediction**: Applied to new store-level data

---

## 2. Business Objective

The MLR model aims to:
- Forecast car seat sales across retail stores  
- Quantify the influence of individual predictors  
- Provide a transparent baseline model for comparison  
- Generate predictions for new stores for operational decisions

---

## 3. Dataset Overview

The dataset contains the following variables:

**Target Variable**  
- **Sales**: Number of car seats sold (in hundreds)

**Predictor Variables**  
- **CompPrice** – Competitor pricing  
- **Income** – Local average income (in $000s)  
- **Advertising** – Advertising spend (in $000s)  
- **Population** – Local population size (in $000s)  
- **Price** – Store price  
- **ShelfLoc** – Shelf location quality (Good, Medium, Bad)  
- **Age** – Average customer age  
- **Education** – Average education level  
- **Urban** – Urban location indicator  
- **US** – U.S. store indicator  

**Excluded Variable**  
- **Store** – Unique store identifier

---

## 4. Environment Setup

```{r message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
library(ggplot2)
```

---

## 5. Data Ingestion

```{r}
car_seat_sales <- read.csv("data/car_seat_sales.csv")
new_car_seat_sales <- read.csv("data/new_car_seat_sales_data.csv")
```

---

## 6. Data Preparation

The `Store` variable is removed as it is a non-predictive identifier.  
Categorical variables are explicitly converted to factors and aligned between training and new data.

```{r}
library(dplyr)

model_data <- car_seat_sales %>%
  select(-Store)

new_model_data <- new_car_seat_sales %>%
  select(-Store)

categorical_vars <- c("ShelfLoc", "Urban", "US")

for (col in categorical_vars) {
  model_data[[col]] <- as.factor(model_data[[col]])
  new_model_data[[col]] <- factor(
    new_model_data[[col]],
    levels = levels(model_data[[col]])
  )
}
```

---

## 7. Modeling Strategy

### Why Multi-Linear Regression?

Multiple Linear Regression provides a transparent and interpretable framework for understanding the relationship between sales and multiple predictors.

Key advantages include:

- Clear interpretation of predictor effects  
- Strong baseline for comparison with non-linear models  
- Useful for explaining sales drivers to business stakeholders  

### Validation Approach

A **5-fold cross-validation** strategy is used to:

- Assess model generalization  
- Produce performance metrics (RMSE, R²)  
- Enable fair comparison with CART, Random Forest, and Neural Networks  

---

## 8. Model Training – Multi-Linear Regression

```{r}
library(caret)

set.seed(1975)

cv_method <- trainControl(
  method = "cv",
  number = 5
)

mlr_model <- train(
  Sales ~ .,
  data = model_data,
  method = "lm",
  trControl = cv_method
)

mlr_model
```

---

## 9. Model Evaluation

```{r}
summary(mlr_model$finalModel)
```

Interpretation guidelines:

- Coefficients indicate the expected change in sales per unit change in predictors  
- **R²** measures the proportion of variance explained  
- **RMSE** represents prediction error magnitude  

---

## 10. Generating Predictions on New Store Data

```{r}
mlr_predictions <- predict(mlr_model, new_model_data)

prediction_results <- new_car_seat_sales %>%
  mutate(Predicted_Sales = mlr_predictions)

head(prediction_results)
```

`Predicted_Sales` represents the expected number of car seats sold,  
measured in **hundreds of units**, for each new store.

---

## 11. Business Insights

Multiple Linear Regression enables clear identification of factors influencing sales:

- **ShelfLoc** and **Price** exhibit the strongest impact on sales  
- Demographic variables such as **Age**, **Income**, and **Education** show smaller but measurable effects  
- The linear baseline helps validate and calibrate more complex models  

---

## 12. Limitations and Next Steps

Limitations of the Multi-Linear Regression model include:

- Assumption of linear relationships between predictors and sales  
- Sensitivity to multicollinearity among predictors  
- Inability to capture complex non-linear interactions  

Planned next steps:

- Regularized regression (Ridge, Lasso, Elastic Net)  
- Cross-model performance comparison  
- Final model selection for deployment  
