---
title: "Car Seat Sales Forecasting Using Neural Networks"
author: "Pranai Basava"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

## Executive Summary

This analysis applies a **Neural Network regression model** to forecast car seat sales using store-level demographic, pricing, and merchandising attributes.

Neural Networks extend tree-based models by learning **complex non-linear relationships** through weighted combinations of input features. Using **5-fold cross-validation**, the model is trained and tuned to balance predictive accuracy and generalization performance.

The trained model is applied to **new, unseen store data** to generate forward-looking sales forecasts that can support inventory planning, pricing strategy, and demand estimation.

This model complements CART and Random Forest approaches by prioritizing **model flexibility and expressive power**, making it well-suited for capturing subtle patterns in retail sales data.

---

## 1. Project Overview

Accurate demand forecasting is essential for optimizing retail operations, including inventory management, pricing decisions, and marketing investments.

This notebook implements a **Neural Network regression model** to forecast car seat sales. The analysis follows the same dataset, validation framework, and evaluation strategy used in other models within this project to ensure **consistent and fair comparison**.

---

## 2. Business Objective

The objectives of this analysis are to:

- Forecast car seat sales across retail locations
- Capture complex, non-linear relationships in the data
- Improve predictive performance beyond tree-based models
- Generate reliable sales estimates for new store configurations

---

## 3. Dataset Overview

The dataset contains store-level attributes related to competition, demographics, pricing, and merchandising.

### Target Variable
- **Sales** – Number of car seats sold (measured in hundreds of units)

### Predictor Variables
- **CompPrice** – Competitor pricing
- **Income** – Local average income (in $000s)
- **Advertising** – Advertising spend (in $000s)
- **Population** – Local population size (in $000s)
- **Price** – Store price
- **ShelfLoc** – Shelf location quality
- **Age** – Average customer age
- **Education** – Average education level
- **Urban** – Urban location indicator
- **US** – U.S. store indicator

The **Store** variable is excluded from modeling as it serves only as a unique identifier.

---

## 4. Environment Setup

```{r message=FALSE, warning=FALSE}
library(caret)
library(dplyr)
```

## 5. Data Ingestion

```{r}
car_seat_sales <- read.csv("data/car_seat_sales.csv")
new_car_seat_sales <- read.csv("data/new_car_seat_sales_data.csv")
```

---

## 6. Data Preparation

The `Store` variable is removed prior to modeling, as it serves only as a unique identifier.  
Neural Networks require numeric input; factor variables are automatically handled by `caret` through internal encoding.


```{r}

model_data <- car_seat_sales %>%
  select(-Store)

new_model_data <- new_car_seat_sales %>%
  select(-Store)
```

---

## 7. Modeling Strategy

### Why Neural Networks?

Neural Networks are capable of learning highly non-linear patterns through weighted connections between input features and hidden layers.

Key advantages include:

- Flexibility in modeling complex relationships  
- Ability to capture subtle feature interactions  
- Strong performance in non-linear forecasting tasks  

### Validation Approach

A **5-fold cross-validation** strategy is employed to:

- Reduce overfitting risk  
- Stabilize performance estimates  
- Enable fair comparison across models  

---

## 8. Hyperparameter Tuning Strategy

Neural Networks require tuning of two key hyperparameters:

- **size** – Number of neurons in the hidden layer  
- **decay** – Regularization parameter to prevent overfitting  

```{r}
set.seed(1975)

cv_method <- trainControl(
  method = "cv",
  number = 5
)

nn_grid <- expand.grid(
  .size = seq(5, 20, 5),
  .decay = seq(0.1, 0.5, 0.1)
)

nn_grid
```

---

## 9. Model Training – Neural Network

```{r}
nn_model <- train(
  Sales ~ .,
  data = model_data,
  method = "nnet",
  trControl = cv_method,
  tuneGrid = nn_grid,
  linout = TRUE,
  trace = FALSE
)

nn_model
```

---

## 10. Hyperparameter Interpretation

- Hidden layer **size** controls model capacity  
- **Decay** penalizes large weights to reduce overfitting  

The optimal combination of `size` and `decay` is selected based on **minimum cross-validated RMSE**, ensuring the model generalizes well to unseen data.

```{r}
nn_model$bestTune
```

---

## 11. Generating Predictions on New Store Data

In a production scenario, the trained Neural Network model is applied to new store-level data without historical sales information.

```{r}
nn_predictions <- predict(nn_model, new_model_data)

prediction_results <- new_car_seat_sales %>%
  mutate(Predicted_Sales = nn_predictions)

head(prediction_results)
```

`Predicted_Sales` represents the expected number of car seats sold,  
measured in **hundreds of units**, for each new store configuration.

---

## 12. Business Insights

Neural Networks provide a flexible modeling framework capable of capturing complex, non-linear demand patterns.

From a business perspective, this enables:

- Improved forecasting in dynamic retail environments  
- Strong performance where linear or tree-based models may underfit  
- Enhanced sensitivity to subtle pricing and demographic effects  

However, Neural Networks trade interpretability for predictive power, making them most suitable when **accuracy is prioritized over transparency**.

---

## 13. Limitations and Next Steps

While Neural Networks offer strong predictive capability, they:

- Require careful hyperparameter tuning  
- Are less interpretable than tree-based models  
- May be sensitive to feature scaling and data distribution  

Future work could include feature engineering, alternative network architectures, and ensemble strategies combining multiple model types.
