---
title: "Car Seat Sales Forecasting Using Random Forest"
author: "Pranai Basava"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

## Executive Summary

This analysis applies a **Random Forest regression model** to forecast car seat sales across retail stores using demographic, pricing, and merchandising attributes.

Random Forest extends the baseline CART model by leveraging an **ensemble of decision trees**, resulting in improved predictive accuracy and robustness to noise and outliers. The model is trained using **5-fold cross-validation** and evaluated using consistent performance metrics to ensure fair comparison across modeling approaches.

The trained model is applied to **new, unseen store data** to generate forward-looking sales forecasts that can support:

- Inventory and demand planning  
- Store-level pricing strategy  
- Advertising budget optimization  
- Risk mitigation for underperforming store configurations  

This Random Forest model serves as a **high-performing benchmark**, balancing predictive accuracy and scalability, while complementing the interpretability of CART and the flexibility of Neural Networks.

---

## 1. Project Overview

Accurate sales forecasting enables retailers to optimize inventory levels, pricing strategies, and marketing investments.

This notebook implements a **Random Forest regression model** to forecast car seat sales using store-level demographic, pricing, and merchandising attributes. The analysis follows the same data pipeline, validation strategy, and evaluation framework as other models in this project to ensure fair and consistent comparison.

---

## 2. Business Objective

The primary objective of this analysis is to:

- Forecast car seat sales across retail stores
- Improve prediction accuracy beyond a single decision tree
- Capture non-linear patterns and feature interactions
- Generate reliable sales estimates for new, unseen stores

---

## 3. Dataset Overview

The dataset contains store-level attributes related to competition, pricing, demographics, and merchandising.

### Target Variable
- **Sales** – Number of car seats sold (measured in hundreds)

### Predictor Variables
- **CompPrice** – Competitor pricing
- **Income** – Local average income (in $000s)
- **Advertising** – Advertising spend (in $000s)
- **Population** – Local population size (in $000s)
- **Price** – Store price
- **ShelfLoc** – Shelf location quality
- **Age** – Average customer age
- **Education** – Average education level
- **Urban** – Urban location indicator
- **US** – U.S. store indicator

The variable **Store** is excluded from modeling as it is a unique identifier.

---

## 4. Environment Setup

```{r message=FALSE, warning=FALSE}
library(caret)
library(randomForest)
library(dplyr)
```

---

## 5. Data Ingestion

The dataset is loaded from the local project directory. This dataset contains historical sales and store-level attributes used for model training.

```{r data-ingestion}
car_seat_sales <- read.csv("data/car_seat_sales.csv")
```

---

## 6. Data Preparation

The `Store` variable is removed from the modeling dataset, as it serves only as a unique identifier and does not contribute to predictive performance.

```{r}
library(dplyr)

model_data <- car_seat_sales %>%
  dplyr::select(-Store)
```

---

## 7. Modeling Strategy

### Why Random Forest?

Random Forest is an ensemble learning method that aggregates predictions from multiple decision trees to improve accuracy and generalization.

Key advantages include:

- Reduced overfitting compared to single trees  
- Improved predictive performance  
- Robust handling of non-linear relationships  
- Implicit modeling of feature interactions  

### Validation Approach

A **5-fold cross-validation** strategy is used to ensure model stability and enable fair comparison with other models in this project.

---

## 8. Hyperparameter Tuning Strategy

The primary hyperparameter in Random Forest regression is **mtry**, which controls the number of candidate features considered at each split.

```{r}
set.seed(1975)

library(caret)

cv_method <- trainControl(
  method = "cv",
  number = 5
)

mtry_grid <- expand.grid(
  .mtry = 1:10
)
```
Tuning the *mtry* parameter allows the model to balance randomness and predictive strength by controlling how many features compete at each split.

---

## 9. Model Training – Random Forest

The Random Forest model is trained using the defined cross-validation strategy and hyperparameter grid.

```{r}
rf_model <- train(
  Sales ~ .,
  data = model_data,
  method = "rf",
  trControl = cv_method,
  tuneGrid = mtry_grid
)

rf_model
```

---

## 10. Model Evaluation

The trained model object contains cross-validated performance metrics and the optimal hyperparameter configuration.

```{r}
rf_model$bestTune
```

---

## 11. Generating Predictions on New Store Data

In a production setting, the trained model is applied to new store configurations where historical sales data is unavailable.

### Load New Data

```{r}
new_car_seat_sales <- read.csv("data/new_car_seat_sales_data.csv")

new_model_data <- new_car_seat_sales %>%
  dplyr::select(-Store)
```

### Generate Predictions

```{r}
rf_predictions <- predict(rf_model, new_model_data)

prediction_results <- new_car_seat_sales %>%
  mutate(Predicted_Sales = rf_predictions)

head(prediction_results)
```

*prediction_results* represents the expected number of car seats sold  
(measured in hundreds of units) for each new store.

---

## 12. Business Insights

The Random Forest model provides improved forecasting accuracy by aggregating multiple decision trees.

This approach delivers:

- More stable predictions  
- Reduced sensitivity to individual data points  
- Strong performance in complex, non-linear environments  

---

## 13. Limitations and Next Steps

Final model selection will be guided by cross-validated performance metrics and business requirements such as interpretability and deployment constraints.



